Spatial distribution of the COVID-19 lockdown crime drop in England and Wales

### 2021-03-22 draft_paper.Rmd
submission docs

---
title: "Spatial distribution of the COVID-19 lockdown crime drop in England and Wales"
author:
  - Samuel Langton
  - Anthony Dixon
  - Graham Farrell
csl: apa.csl
header-includes: \usepackage{float}
output:
  word_document:
    reference_docx: reference_doc.docx
  pdf_document: default
bibliography: refs.bib
---

@@

@@

Recent months have seen a wave of interest in describing and explaining how crime has changed in response to restrictions on mobility brought about to limit the spread of COVID-19. Studies have been undertaken in the United States [@campedelli2020exploring; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show], Sweden [gerell2020minor], China (@borrion2020measuring) and Mexico [@estevez2020crime]. Generally speaking, findings have been consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@halford2020crime]. That is, legally-enforced restrictions on daily activities, mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime. As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often back to levels that would have been expected without the global pandemic and ensuing lockdowns [@langton2021cs]

@@

Open police-recorded data on crime and anti-social behavior in England and Wales is published through an online web portal (https://data.police.uk/). Individual records are released on a month-by-month basis for each of the 43 police forces comprising England and Wales. We used a study period spanning February to August 2020 in order to capture the first six months of the nationwide lockdown (March to August) and the one month preceding the change (February). For reference and comparison to historical trends, we obtained data for the same months in 2018 and 2019. Individual records are time-stamped by month -- the temporal scale of this study. Due to incomplete data releases from Greater Manchester Police, we excluded data from the Greater Manchester region, collating data from 42 out of 43 forces in England and Wales.

@@

To summarize the opportunity structure of local areas we required a nationwide dataset of theoretically relevant facilities and urban features which could be aggregated at the LSOA level. To this end, we obtained point-level data from the Application Programming Interface (API) for Open Street Map via the `osmdata` package [@padgham2017osmdata] in `R` [@team2013r]. Open Street Map is a crowdsourced geospatial database containing a vast array of features which can be used for explaining the temporal and spatial patterning of crime [@langton2020open]. Geographic features are identified by pairs of keys and values which can be used to computationally query the API for geospatial data. Based on existing research examining the opportunity structures of fine-grained spatial scales, we collated the coordinate locations of the following facilities:

@@

Macro-level descriptives of count trends notifiable offences (excluding drugs) are reported for provide context from which we will subsequently unmask local (meso) variation. We visualize observed counts between February and August 2020 relative to the same periods in 2018 and 2019. In doing so, we can observe how nationwide crime trends changed in the face of lockdown measures [see also @langton2021cs]. This sets the scene from which we can disentangle the underlying meso-level variation.

@@

We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offence (excluding drug) counts for LSOA in England and Wales (N = 33,075) between February and August 2020. To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite]. For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable. Potential cluster solution options were also examined using Principal Component Analysis to establish their suitability.

@@

To set the context for the localized analysis, Figure 2 visualizes crime counts between February and August 2020 for notifiable offences excluding drugs before and after lockdown. This descriptive analysis broadly confirms what has been demonstrated using model-based forecasting analyses elsewhere (Langton et al., 2021) although we note that this descriptive comparison does not account for long-term trends. In April, the first full month of lockdown in England and Wales, we observe a nationwide decline in notifiable offences in comparison to previous years. Upon the relaxation of lockdown rules, crime resurged back to normality, and by August, crime had returned to within a range we might have expected without the nationwide lockdown. This trend represents the 'global' trend which we will subsequently disentangle using localized analyses.

```{r nocounts, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Notitifable offence (excluding drugs) counts in England and Wales during the lockdown period.", fig.align='center'}
knitr::include_graphics("visuals/no_raw_counts_gg.png")
```

@@

We deploy non-parametric k-means clustering on LSOA (meso) level geographic units (N = 33,075) comprising England and Wales to decompose the macro-level 'decline and resurgence' observed between February and August, 2020. Based on an assessment of the Calinski-Criterion statistic [@calinski1974dendrite] and Principal Component Analysis (see Appendix), we determined the optimal solution to be 6 clusters (see Figure 3). Solid black lines represent the median count for each cluster at any given time point, with the black dotted line showing the mean. Violin plots have been added to convey variation around these points at each time point. These suggest that the mean and median point statistics summarize the underlying data reasonably well, and indicate that clusters are distinct from one another, with little overlap. Additional lines have been added to convey each clusters' mean and median trend in 2018 and 2019 respectively. These trends suggest that the clusters identified using the 2020 study period were distinct and meaningful even in previous years, and provide a relative baseline from which we can compare lockdown trends.

Overall, we note that most LSOA were remarkably stable during the pandemic. Clusters A and B could be described and 'low crime and stable', exhibiting fairly low counts throughout the study period and across years. Together, these clusters comprise 89% of LSOA in England and Wales. Even amidst the stark macro-level decline in notifiable offences (see Figure 2) LSOA in these clusters only experienced marginal average dips in crime. The third largest cluster, cluster C, comprises 10% of LSOA in England and Wales. LSOA in this cluster experienced a more prominent dip in crime, along with cluster D (1.6% of LSOA). Together, we might describe these LSOAs as 'mid-crime, mid-drop'. In both cases, there is a clear deviation from previous years. Notifiable offences fall between March and April, and then began to converge back to levels observed in previous years.

@@

@@

To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the monthly change in counts and percentage of total absolute change (i.e. positive or negative) attributable to each cluster. We can use this visual to identify which clusters drove the initial decline and subsequent nationwide resurgence in crime.

As expected, the vast majority of change across all clusters occurred between March and April. Between these months, notifiable offences experienced a dramatic fall nationwide. That said, the figure demonstrates that this decline did not occur equitably across local areas. Consistent with the cluster solutions trends in Figure 3, a small number of LSOA (meso) units appear to have contributed disproportionately to the nationwide (macro) trend. For instance, clusters A and B, which comprise 58% and 31% of LSOA in the country, accounted for only 20% and 27% of the total decline between March and April. By contrast, clusters E and F, which together comprised only 0.33% of LSOA (N = 110) contributed to 15% of the nationwide drop in these months. Again, this suggests that most areas remarkably stable, given the nationwide volatility, with just a handful of areas driving the decline the nationwide lockdown crime drop.

@@

The 'low crime and stable' clusters (A and B) actually _declined again_ between July and August, and the higher crime clusters continued to increase. The change is almost imperceptible when viewing average counts in these LSOA (see Figure 3) but the sheer size of these units, collectively comprising 89% of England and Wales, has a major impact on nationwide trends. This is suggestive of highly localized change in opportunities as lockdown rules were relaxed: change which is aggregated away at the macro-level. Stable, low crime areas mitigated against further nationwide increases which may, if they had continued, resulted in a higher crime count in August than observed in previous years.

@@

These findings can be supplemented with a comparison to the previous year. Figure 5 compares the clusters identified in 2020 with the equivalent months in 2019. Here, it is clear that the 'high crime major drop' clusters (E and F) contributed the most to the lockdown crime drop relative to their size. By way of example, around 27,000 fewer crimes occurred in cluster A--containing 19,162 LSOA--during April 2020 compared to April 2019 (Figure 5a). This translates to a 24% drop (Figure 5b), and an average fall per LSOA of 1.39 counts (Figure 5c). For cluster F, containing just 9 LSOA, approximately 5,000 fewer crimes occurred (Figure 5a), which translates to an 87% drop (Figure 5b), and an average fall per LSOA of 543 counts (Figure 5c). In other words: whether comparing month-to-month change in 2020, or comparing equivalent months in 2019, the lockdown crime drop was disproportionately driven by a small number of local areas, with most meso-level units remaining fairly stable, even amidst dramatic macro-level change.

@@

@@

@@

@@

```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type characteristics of each cluster solution.", out.width= "100%", fig.align='center'}
knitr::include_graphics("visuals/traj_crimes_monthly_k6_gg.png")
```

@@

This paper has aimed to investigate the localized drivers of the COVID-19 crime drop in England and Wales. Using 7-months of police-recorded data on notifiable offences (excluding drugs), we decomposed the (macro) nationwide decline in crime according to clusters of (meso) local areas, each with distinct longitudinal trajectories. Findings provide substantial evidence to suggest that the nationwide decline in crime during lockdown has been driven by a disproportionately small number of localized geographic areas. We find that these localized drivers, dominating city centers, had pre-existing high levels of crime, likely attributable to their unique opportunity structures, containing numerous nightlife facilities, commercial shops, public transport nodes and bicycle parking spaces. Clusters demonstrate a degree of stability in their crime type profiles, but the 'high crime major drop' areas appear to be weighted towards theft and shoplifting. Despite their dramatic declines, these areas experienced a fairly slow resurgence upon the relaxation of lockdown rules, remaining below historical levels even by the time low and mid-crime clusters had returned to normality.

@@

```{r crimegini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Monthly generalized Gini cofficient trends by offence type in England and Wales during the lockdown period.", fig.align='center'}
knitr::include_graphics("visuals/gini_gg.png")
```

>>>>

---
title: "Spatial distribution of the COVID-19 lockdown crime drop in England and Wales"
author:
- Samuel Langton (University of Leeds)
- Anthony Dixon (University of Leeds) 
- Graham Farrell (University of Leeds)
csl: apa.csl
output:
  word_document:
    reference_docx: reference_doc.docx
header-includes: \usepackage{float}
bibliography: refs.bib
---

@@

**Abstract**

Studies from international case study sites have documented the impact of the COVID-19 pandemic and ensuing lockdown policies on crime. Restrictions on mobility and social interaction have brought about widespread falls in police-recorded crime, swiftly followed by a resurgence as lockdowns were lifted. Less is known about the localized spatial patterning of these macro-level trends. This study uses police-recorded data on notifiable offences to decompose the nationwide lockdown trend in England and Wales. We find that local, meso-level areas did not experience the decline and subsequent 'bounce back' in unison. Rather, most areas remained remarkably stable, with a small number of meso units contributing a disproportionate amount to the macro-level trend. We find that these areas often occupy city centers, had plentiful opportunities for crime pre-pandemic and were dominated by theft and shoplifting offences.

**Keywords:** COVID-19, clustering, kmeans, routine activities, opportunity theory, pandemic.

**Corresponding author:** Graham Farrell, School of Law, University of Leeds, United Kingdom.

Email: g.farrell@leeds.ac.uk.

**Competing interests:** The authors declare that they have no competing interests.

**Availability of data and materials:** The data are publicly available and retrievable from the open online data portal for England, Wales and Northern Ireland. Code to replicate data downloads, handling and analyses is openly available (https://anonymous.4open.science/r/46199882-cc5a-449d-9946-300b3132ab2e/).

\newpage

@@

Recent months have seen a wave of interest in describing and explaining how crime has changed in response to restrictions on mobility brought about to limit the spread of COVID-19. Studies have been undertaken in the United States [@campedelli2020exploring; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show], Sweden [@gerell2020minor], China [@borrion2020measuring] and Mexico [@estevez2020crime]. Generally speaking, findings have been consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@halford2020crime]. That is, legally-enforced restrictions on daily activities, mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime. As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often back to levels that would have been expected without the global pandemic and ensuing lockdowns [@langton2021cs].

@@

Open police-recorded data on crime and anti-social behavior in England and Wales is published through an online web portal (https://data.police.uk/). Individual records are released on a month-by-month basis for each of the 43 police forces comprising England and Wales. We used a study period spanning February to August 2020 in order to capture the first six months of the nationwide lockdown (March to August) and the one month preceding the change (February). Here, we note that lockdown was initiated on 23 March, making April the first full month of measures. For reference and comparison to historical trends, we obtained data for the same months in 2018 and 2019. Individual records are time-stamped by month -- the temporal scale of this study. Due to incomplete data releases from Greater Manchester Police, we excluded data from the Greater Manchester region, collating data from 42 out of 43 forces in England and Wales.

@@

To summarize the opportunity structure of local areas we required a nationwide dataset of theoretically relevant facilities and urban features which could be aggregated at the LSOA level. To this end, we obtained point-level data from the Application Programming Interface (API) for Open Street Map via the _osmdata_ package [@padgham2017osmdata] in _R_ [@team2013r]. Open Street Map is a crowdsourced geospatial database containing a vast array of features which can be used for explaining the temporal and spatial patterning of crime [@langton2020open]. Geographic features are identified by pairs of keys and values which can be used to computationally query the API for geospatial data. Based on existing research examining the opportunity structures of fine-grained spatial scales, we collated the coordinate locations of the following facilities:

@@

Macro-level descriptives of count trends notifiable offences (excluding drugs) are reported for provide context from which we will subsequently unmask local (meso) variation. We visualize observed counts between February and August 2020 relative to the same periods in 2018 and 2019. In doing so, we can observe how crime trends changed in the face of lockdown measures in England and Wales [see also @langton2021cs]. This sets the scene from which we can disentangle the underlying meso-level variation.

@@

We deploy k-means using the _kml_ package [@kmlpack] in R on notifiable offence (excluding drug) counts for LSOA in England and Wales (N = 33,075) between February and August 2020. To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite]. For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable. Potential cluster solution options were also examined using Principal Component Analysis to establish their suitability.

@@

To set the context for the localized analysis, Figure 1 visualizes crime counts between February and August 2020 for notifiable offences excluding drugs before and after lockdown. In April, the first full month of lockdown in England and Wales, we observe a nationwide decline in notifiable offences in comparison to previous years. Upon the relaxation of lockdown rules, crime resurged back to normality, and by August, crime had returned to within a range we might have expected without the nationwide lockdown [see also @langton2021cs]. This trend represents the 'global' trend which we will subsequently disentangle using localized analyses.

```{r nocounts, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Notifiable offence (excluding drugs) end of month counts in England and Wales. April was the first full month of lockdown.", fig.align='center'}
knitr::include_graphics("visuals/no_raw_counts_gg.png")
```

[Insert Figure 1 here. Caption: Notifiable offence (excluding drugs) end of month counts in England and Wales. April was the first full month of lockdown.]

@@

We deploy non-parametric k-means clustering on LSOA (meso) level geographic units (N = 33,075) comprising England and Wales to decompose the macro-level 'decline and resurgence' observed between February and August, 2020. Based on an assessment of the Calinski-Criterion statistic [@calinski1974dendrite] and Principal Component Analysis, we determined the optimal solution to be 6 clusters (see Figure 2). Solid black lines represent the median count for each cluster at any given time point, with the black dotted line showing the mean. Violin plots have been added to convey variation around these points at each time point. These suggest that the mean and median point statistics summarize the underlying data reasonably well, and indicate that clusters are distinct from one another, with little overlap. Additional lines have been added to convey each clusters' mean and median trend in 2018 and 2019 respectively. These trends suggest that the clusters identified using the 2020 study period were distinct and meaningful even in previous years, and provide a relative baseline from which we can compare lockdown trends.

Overall, we note that most LSOA were remarkably stable during the pandemic. Clusters A and B could be described and 'low crime and stable', exhibiting fairly low counts throughout the study period and across years. Together, these clusters comprise 89% of LSOA in England and Wales. Even amidst the stark macro-level decline in notifiable offences (see Figure 1) LSOA in these clusters only experienced marginal average dips in crime. The third largest cluster, cluster C, comprises 10% of LSOA in England and Wales. LSOA in this cluster experienced a more prominent dip in crime, along with cluster D (1.6% of LSOA). Together, we might describe these LSOAs as 'mid-crime, mid-drop'. In both cases, there is a clear deviation from previous years. Notifiable offences fall between March and April, and then began to converge back to levels observed in previous years.

@@

[Insert Figure 2 here. Caption: K-means cluster solutions for LSOA notifiable offences (excluding drugs). Distributions refer to 2020 only.]

@@

To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 3 plots the monthly change in counts and percentage of total absolute change (i.e. positive or negative) attributable to each cluster. We can use this visual to identify which clusters drove the initial decline and subsequent nationwide resurgence in crime.

As expected, the vast majority of change across all clusters occurred between March and April. Between these months, notifiable offences experienced a dramatic fall nationwide. That said, the figure demonstrates that this decline did not occur equitably across local areas. Consistent with the cluster solutions trends in Figure 2, a small number of LSOA (meso) units appear to have contributed disproportionately to the nationwide (macro) trend. For instance, clusters A and B, which comprise 58% and 31% of LSOA in the country, accounted for only 20% and 27% of the total decline between March and April. By contrast, clusters E and F, which together comprised only 0.33% of LSOA (N = 110) contributed to 15% of the nationwide drop in these months. Again, this suggests that most areas remarkably stable, given the nationwide volatility, with just a handful of areas driving the decline the nationwide lockdown crime drop.

@@

The 'low crime and stable' clusters (A and B) actually _declined again_ between July and August, and the higher crime clusters continued to increase. The change is almost imperceptible when viewing average counts in these LSOA (see Figure 2) but the sheer size of these units, collectively comprising 89% of England and Wales, has a major impact on nationwide trends. This is suggestive of highly localized change in opportunities as lockdown rules were relaxed: change which is aggregated away at the macro-level. Stable, low crime areas mitigated against further nationwide increases which may, if they had continued, resulted in a higher crime count in August than observed in previous years.

@@

[Insert Figure 3 here. Caption: Counts and percentage of nationwide change between months attributable to each cluster.]

These findings can be supplemented with a comparison to the previous year. Figure 4 compares the clusters identified in 2020 with the equivalent months in 2019. Here, it is clear that the 'high crime major drop' clusters (E and F) contributed the most to the lockdown crime drop relative to their size. By way of example, around 27,000 fewer crimes occurred in cluster A--containing 19,162 LSOA--during April 2020 compared to April 2019 (Figure 4a). This translates to a 24% drop (Figure 4b), and an average fall per LSOA of 1.39 counts (Figure 4c). For cluster F, containing just 9 LSOA, approximately 5,000 fewer crimes occurred (Figure 4a), which translates to an 87% drop (Figure 4b), and an average fall per LSOA of 543 counts (Figure 4c). In other words: whether comparing month-to-month change in 2020, or comparing equivalent months in 2019, the lockdown crime drop was disproportionately driven by a small number of local areas, with most meso-level units remaining fairly stable, even amidst dramatic macro-level change.

@@

[Insert Figure 4 here. Caption: Comparing LSOA clusters between equivalent months in 2020 and 2019 according to (a) absolute difference in counts, (b) percentage difference in counts, (c) average difference in counts.]

@@

[Insert Figure 5 here. Caption: Spatial distribution of clusters by major urban conurbation. LSOA with the highest count of Open Street Map features (in black) denotes city centre.]

@@

[Insert Table 1 here. Caption: Descriptive statistics of facilities and features in each cluster. Sourced from Open Street Map.]

@@

```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type profiles of each cluster solution.", out.width= "100%", fig.align='center'}
knitr::include_graphics("visuals/traj_crimes_monthly_k6_gg.png")
```

[Insert Figure 6 here. Caption: Crime type profiles of each cluster solution.]

@@

This paper has aimed to investigate the localized drivers of the COVID-19 crime drop in England and Wales. Using 7-months of police-recorded data on notifiable offences (excluding drugs), we decomposed the (macro) nationwide decline in crime according to clusters of (meso) local areas, each with distinct longitudinal trajectories. Findings provide substantial evidence to suggest that the nationwide decline in crime during lockdown has been driven by a disproportionately small number of localized geographic areas. We find that these localized drivers, dominating city centers, had pre-existing high levels of crime, likely attributable to their unique opportunity structures, containing numerous nightlife facilities, commercial shops, public transport nodes and bicycle parking spaces. Clusters demonstrate a degree of stability in their crime type profiles, but the 'high crime, major drop' areas appear to be weighted towards theft and shoplifting. Despite their dramatic declines, these areas experienced a fairly shallow resurgence upon the relaxation of lockdown rules, remaining below historical levels even by the time low and mid-crime clusters had returned to normality.

@@

[Insert Figure 7 here. Caption: Counts by offence type in England and Wales during the lockdown period.]

```{r crimegini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Monthly generalized Gini coefficient trends by offence type in England and Wales during the lockdown period.", fig.align='center'}
knitr::include_graphics("visuals/gini_gg.png")
```

[Insert Figure 8 here. Caption: Monthly generalized Gini coefficient trends by offence type in England and Wales during the lockdown period.]
